
\documentclass[a4paper,11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{dirtree}

\renewcommand\familydefault{\sfdefault}
\usepackage{tgheros}

\usepackage{amsmath,amssymb,amsthm,textcomp}
\usepackage{enumerate}
\usepackage{multicol}
\usepackage{tikz}

\usepackage{geometry}
\geometry{left=25mm,right=25mm,%
bindingoffset=0mm, top=20mm,bottom=20mm}


\linespread{1.3}

\newcommand{\linia}{\rule{\linewidth}{0.5pt}}

% custom theorems if needed
\newtheoremstyle{mytheor}
    {1ex}{1ex}{\normalfont}{0pt}{\scshape}{.}{1ex}
    {{\thmname{#1 }}{\thmnumber{#2}}{\thmnote{ (#3)}}}

\theoremstyle{mytheor}
\newtheorem{defi}{Definition}

% my own titles
\makeatletter
\renewcommand{\maketitle}{
\begin{center}
\vspace{2ex}
{\huge \textsc{\@title}}
\vspace{1ex}
\\
\linia\\
\@author \hfill \@date
\vspace{4ex}
\end{center}
}
\makeatother
%%%

% custom footers and headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\chead{}
\rhead{}
\lfoot{Learning Python with Doudou}
\cfoot{}
\rfoot{Page \thepage}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
%

% code listing settings
\usepackage{listings}
\lstset{
    language=Python,
    basicstyle=\ttfamily\small,
    aboveskip={1.0\baselineskip},
    belowskip={1.0\baselineskip},
    columns=fixed,
    extendedchars=true,
    breaklines=true,
    tabsize=4,
    prebreak=\raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
    frame=lines,
    showtabs=false,
    showspaces=false,
    showstringspaces=false,
    keywordstyle=\color[rgb]{0.627,0.126,0.941},
    commentstyle=\color[rgb]{0.133,0.545,0.133},
    stringstyle=\color[rgb]{01,0,0},
    numbers=left,
    numberstyle=\small,
    stepnumber=1,
    numbersep=10pt,
    captionpos=t,
    escapeinside={\%*}{*)}
}

%%%----------%%%----------%%%----------%%%----------%%%

\begin{document}

\title{Learning Python with Doudou}

\author{Brian Teixeira, Doudou University}

\date{05/25/2020}

\maketitle

Welcome to this Python class! In this series of tutorials, you will
learn how to fully master the python programming language through
explanations, examples and exercises. This python class will be split
into three main parts: Basics, Intermediate and Advanced. Following
this tutorial, you will be able to unlock a new series of tutorials
called Machine Learning with Python! Pretty exciting right?

\vspace{5mm}

The ``Basics`` will mostly present general stuff about the python
programming language. You will get to learn about functions,
variables, scopes, packages etc. Some of the things will sound very
basic and obvious, but it's always great to take a fresh look at things.
I will try to go as far as I can while describing the behavior of the
language, especially with regards to the management of the memory.

\vspace{5mm}

The ``Intermediate`` part will be the longest part of this tutorial,
not because it's the most complex, but because it's the one that
will cover the biggest part of the language, because as you will understand,
the python language, while rich, is not complex. In this section, you will
get to learn about objects, exceptions, generators and other stuff. I will
insist a lot on the object oriented programming part, as this is, in my
opinion, a very important aspect of the language, and of programming in general.
Don't make me say everything should always be object, but in Python it's kind
of the case. I love objects, but only when things are well designed. Thus,
in this part I will go a bit further than just focusing on the language itself
and will go through some good design practices.

\vspace{5mm}

Finally, the ``Advanced`` part will go through some more rarely used features
of the language, including multiprocessing or networking. You will quickly
realize that this part is not called advanced because it's difficult, but rather
because it's about things you will be less likely to use in your everyday
python journey.

\vspace{5mm}

For each point of the lessons, you will find definitions, examples and
exercises.  For the first lessons, the exercises will be very fast and
academic, but the more you progress in this series of tutorials, the
more they will turn into projects.  There is no recommended time to
finish exercises or projects, because we do not code against the
clock. I will create repositories on github where you will be able
to push the code. Just let me know when you are done with an exercise
or a project, and I will correct it!

\vspace{5mm}

All right, enough talking, let's go!

\newpage

\tableofcontents

\newpage

\include{basics}

\end{document}
